---
title: "Cross-study transcriptomic meta-analyses: selected case studies"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
    code-fold: true
    fig-cap-location: bottom
params:
  layer_b_dir: ""
execute:
  echo: false
  warning: false
  message: false
---

```{=html}
<style>
/* Le tabelle dei valori per gene hanno otto colonne e, a larghezza piena,
   uscivano dal contenitore del testo sovrapponendosi all'indice laterale.
   Regola: la tabella non e' mai piu' larga della colonna di testo, e se il
   contenuto non ci sta scorre DENTRO il proprio riquadro -- la pagina non
   scorre mai in orizzontale. */
.tabella-valori {
  overflow-x: auto;
  max-width: 100%;
  margin-bottom: 1.5rem;
}
.tabella-valori table {
  font-size: 0.82rem;
  margin-bottom: 0;
  width: auto;
  max-width: 100%;
}
.tabella-valori th, .tabella-valori td {
  padding: 0.2rem 0.5rem;
  white-space: nowrap;
}
.tabella-valori caption {
  white-space: normal;
  font-size: 0.85rem;
  caption-side: top;
  text-align: left;
}
</style>
```

```{r setup}
library(simulomicsr)
lb <- load_layer_b(params$layer_b_dir)

# La sezione conclusiva sul corpus completo legge il deliverable annotato dal
# path registrato in run_metadata: i numeri vengono dal file, mai ricopiati a
# mano. Si calcola QUI, in testa, perche' l'introduzione dichiara quante
# meta-analisi ci sono in tutto e da quante e' estratta la selezione: due
# letture dello stesso file potrebbero divergere.
corpus <- simulomicsr:::.corpus_overview_from_bundle(lb)
n_totali_txt <- if (isTRUE(corpus$disponibile) && !is.null(corpus$tabelle)) {
  as.character(corpus$tabelle$n_gruppi)
} else {
  NA_character_
}
```

## Introduction

This document presents `r nrow(lb$selection_resolved)` cross-study transcriptomic
meta-analyses in full. Each one pools differential expression estimates from
independent public studies that measure the same experimental contrast — a
treatment against its own control, or disease tissue against healthy tissue —
under a random-effects model.

The `r nrow(lb$selection_resolved)` were chosen to cover the situations a reader
needs in order to judge the approach: a receptor agonist and its antagonist,
assembled separately and never compared during their construction; a
high-powered cytokine response; a pathogen; an interferon; a small-molecule
inhibitor; a disease; and two groups whose stated limitations are part of what
they are meant to show. `r if (!is.na(n_totali_txt)) sprintf("They are drawn from a set of %s meta-analyses, described in full in the final section.", n_totali_txt) else ""`

For each case study, in this order: a summary card stating what was compared, on
how many studies, and how much they agree; a narrative placing the result in its
biological context, with references; the figures; and the underlying per-gene
values.

```{r case-studies, results = "asis"}
# Le immagini si referenziano con path relative (l'identificativo del gruppo
# seguito dal nome del file, es. "930c8dcf/volcano.png") perche' execute_dir
# del render e' impostato a params$layer_b_dir: Pandoc le trova e le
# base64-embedda (embed-resources: true). Una path assoluta farebbe scattare
# l'auto-prepend "./" di Pandoc, generando un riferimento non risolvibile.
#
# Le figure escluse dal bundle (config$figure_escluse, di default "ma" e
# "heterogeneity") si leggono dal run_metadata invece di essere elencate a
# mano qui: un elenco cablato e' esattamente il difetto gia' corretto una
# volta nello stub di narrative.qmd.
figure_escluse <- unlist(lb$run_metadata$config$figure_escluse)
if (is.null(figure_escluse)) figure_escluse <- character(0)
candidati_figura <- setdiff(
  c("volcano", "forest", "heatmap", "go_enrichment", "ma", "heterogeneity"),
  figure_escluse
)

# Le colonne della tabella per gene, rinominate per la lettura. `gene_label` e
# `gene_id` non compaiono: il primo duplica `gene_symbol`, il secondo (l'ID
# Ensembl) resta come ultima colonna perche' serve a chi verifica.
COLONNE_VALORI <- c(
  gene_symbol           = "Gene",
  logFC_pool            = "log2 FC",
  FDR_BH_within_cluster = "FDR",
  k_effective           = "Studies",
  SE_pool               = "SE",
  tau2                  = "tau-squared",
  I2                    = "I-squared (%)",
  gene_id               = "Ensembl ID"
)

n_caso <- 0L
for (cl_id in names(lb$cluster_bundles)) {
  n_caso <- n_caso + 1L
  bundle <- lb$cluster_bundles[[cl_id]]
  label_paper <- lb$selection_resolved$label_paper[lb$selection_resolved$cluster_id == cl_id]

  # Titolo numerato (richiesta utente 2026-08-06), con la SOLA etichetta
  # leggibile: mai l'identificativo interno del gruppo, che vive nel blocco
  # PROVENANCE della scheda.
  cat(sprintf("\n\n## Case study %d — %s\n\n", n_caso, label_paper))

  # Scheda + narrativa: gia' un unico file per costruzione (narrative.qmd,
  # vedi .write_narrative_template() in R/layer-b-summary-card.R). Qui si
  # toglie il titolo che il file porta gia' (ripeterebbe il ## appena
  # stampato) e si abbassano le sue intestazioni di un livello, cosi' restano
  # annidate sotto il case study.
  narr_path <- file.path(bundle$cluster_dir, "narrative.qmd")
  if (file.exists(narr_path)) {
    narr_lines <- readLines(narr_path)
    body_start <- which(narr_lines == "---")
    if (length(body_start) >= 2L) {
      narr_lines <- narr_lines[(body_start[2L] + 1L):length(narr_lines)]
    }
    narr_lines <- narr_lines[!grepl("^#\\s", narr_lines)]
    # Rete di sicurezza esplicita: nessuna intestazione, in nessun ramo, puo'
    # portare l'identificativo interno di QUESTO gruppo.
    narr_lines <- narr_lines[!(grepl("^#+\\s", narr_lines) & grepl(cl_id, narr_lines, fixed = TRUE))]
    narr_lines <- sub("^## ", "### ", narr_lines)
    cat(paste(narr_lines, collapse = "\n"), "\n\n")
  }

  # Le figure vere (PNG), dopo scheda e narrativa.
  for (plot_name in candidati_figura) {
    abs_path <- file.path(bundle$cluster_dir, sprintf("%s.png", plot_name))
    rel_path <- sprintf("%s/%s.png", cl_id, plot_name)
    caption <- bundle$captions[[plot_name]] %||% sprintf("%s (no caption)", plot_name)
    if (file.exists(abs_path)) {
      cat(sprintf("\n![%s](%s)\n\n", caption, rel_path))
    }
  }

  # I valori per gene dietro le figure. La didascalia di
  # .build_top_gene_table() dichiara il filtro di copertura applicato (es.
  # "Genes measured in fewer than 30 of 59 studies were excluded from
  # ranking") -- senza passarla a kable() il lettore vedrebbe la tabella senza
  # sapere che dei geni sono stati esclusi dall'ordinamento.
  tg_csv <- file.path(bundle$cluster_dir, "top_genes.csv")
  if (file.exists(tg_csv) && file.info(tg_csv)$size > 0L) {
    df <- readr::read_csv(tg_csv, show_col_types = FALSE)
    if (nrow(df) > 0L) {
      cat("\n### Top-ranked genes\n\n")
      colonne <- intersect(names(COLONNE_VALORI), names(df))
      df_vista <- df[, colonne, drop = FALSE]
      names(df_vista) <- unname(COLONNE_VALORI[colonne])
      caption_tg <- bundle$captions$top_gene_table %||%
        "Top differentially expressed genes."
      cat("\n<div class=\"tabella-valori\">\n")
      print(knitr::kable(df_vista, format = "html", digits = 3, caption = caption_tg))
      cat("\n</div>\n\n")
    }
  }
}
```

## The full set of meta-analyses

The case studies above are a selection. This section describes the whole set
from which they are drawn, so that the selection can be judged against it.

```{r corpus-figura, results = "asis"}
# La figura e le tabelle si generano QUI, al momento del render, leggendo il
# deliverable annotato dal path registrato in run_metadata: i numeri vengono
# dal file, mai ricopiati a mano -- un futuro re-render riflette sempre il
# deliverable vero. Se il file non e' raggiungibile da questo render (disco
# esterno non montato, bundle spostato, ...) si dichiara il motivo invece di
# lasciare un buco silenzioso.
if (isTRUE(corpus$disponibile)) {
  cat(sprintf("\n![%s](%s)\n\n",
              corpus$plot$caption, basename(corpus$plot$png_path)))
} else {
  cat(sprintf(paste0(
    "\n::: {.callout-important}\n",
    "This section could not be computed for this render: %s\n:::\n\n"
  ), corpus$motivo))
}
```

```{r corpus-tabelle, results = "asis"}
tb <- corpus$tabelle
if (!is.null(tb)) {
  .n <- function(x) if (is.na(x)) "not available" else format(x, big.mark = ",")
  .pc <- function(x, tot) if (is.na(x)) "" else sprintf(" (%.0f%%)", 100 * x / tot)

  cat(sprintf(paste0(
    "The set comprises **%s meta-analyses**, together reporting %s genes ",
    "significant at FDR < 0.05. The median number of pooled studies is %s and ",
    "the median heterogeneity is I-squared = %.1f%%.\n\n"),
    .n(tb$n_gruppi), .n(tb$totale_geni_sig), .n(tb$mediana_k), tb$mediana_i2))

  cat(sprintf(paste0(
    "Two properties are reported for the whole set because they bound what ",
    "pooling can deliver. **%s%s of the meta-analyses are dominated by a ",
    "single study**, which carries at least half of the inverse-variance ",
    "weight, and **%s%s are worth fewer than two effective studies** by Kish's ",
    "measure. This is not a defect of the estimator — inverse-variance ",
    "weighting must behave this way when one study is far more precise than ",
    "the others — but for those groups pooling adds little beyond the ",
    "heaviest contribution, and the number is stated here rather than left to ",
    "be discovered per group. A further %s%s pool in vitro models together ",
    "with patient-derived tissue.\n\n"),
    .n(tb$n_dominati), .pc(tb$n_dominati, tb$n_gruppi),
    .n(tb$n_sotto_due_efficaci), .pc(tb$n_sotto_due_efficaci, tb$n_gruppi),
    .n(tb$n_materiale_misto), .pc(tb$n_materiale_misto, tb$n_gruppi)))

  cat("\n### Composition by entity type\n\n")
  cat("\n<div class=\"tabella-valori\">\n")
  print(knitr::kable(tb$composizione, format = "html", digits = 1,
                     caption = paste0(
                       "Entity type is read from the prefix of the contrast ",
                       "identifier (ChEBI, HGNC, NCBI Taxonomy, MeSH). ",
                       "Identifiers that were never resolved to an ontology ",
                       "are reported as such rather than assigned to a class.")))
  cat("\n</div>\n\n")

  cat("\n### Power and single-study leverage, by group size\n\n")
  cat("\n<div class=\"tabella-valori\">\n")
  print(knitr::kable(tb$potenza, format = "html", digits = 1,
                     caption = paste0(
                       "The effective number of studies is Kish's measure, ",
                       "(sum of weights)squared / sum of squared weights, on ",
                       "the same inverse-variance weights used for pooling. ",
                       "Single-study dominance is concentrated in the smaller ",
                       "groups.")))
  cat("\n</div>\n\n")

  cat("\n### The meta-analyses with the most pooled studies\n\n")
  cat("\n<div class=\"tabella-valori\">\n")
  print(knitr::kable(tb$piu_potenti, format = "html", digits = 1,
                     caption = paste0(
                       "Ordered by number of pooled studies. Significant ",
                       "genes are counted at FDR < 0.05 within each ",
                       "meta-analysis.")))
  cat("\n</div>\n\n")

  if (nrow(tb$incoerenti) > 0L) {
    cat(sprintf("\n### Meta-analyses declared incoherent (%d)\n\n", nrow(tb$incoerenti)))
    cat(paste0(
      "These groups are reported rather than removed. In each of them at least ",
      "one pooled comparison does not measure the contrast the group is named ",
      "for; the reason is stated so that the estimate is read with it. One of ",
      "them is included among the case studies above.\n\n"))
    cat("\n<div class=\"tabella-valori\">\n")
    print(knitr::kable(tb$incoerenti, format = "html",
                       caption = "Groups whose pooled comparisons were found not to measure a single contrast."))
    cat("\n</div>\n\n")
  }
}
```

## Appendix

### How comparisons with a design defect were counted

For every meta-analysis, the **whole** labels — not truncated: the length
distribution was measured and the suspicious peaks inspected one by one — of the
comparisons that were **actually pooled**, rather than merely assembled, were
extracted. A first reader per block of ten groups gave a verdict against a
rubric of nine known mechanisms (clinical against experimental material, an
entity that is a class rather than a molecule, an entity that is a reagent, a
control that is not a control, incompatible materials, an uncaptured
combination, the wrong entity, arms unmatched for time, donor, cell line,
culture passage or anatomical site); a second reader per block was asked to
refute the first. For the groups large enough to carry a figure, a third agent
counted, comparison by comparison, which ones are defective.

**Where a label admitted two readings** — one under which the comparison is
defective and one under which it is not — the comparison was counted as
**defective**. Ten comparisons were resolved this way: an arm labelled only as
inflamed, in a study that keeps its disease-labelled arms separate; a treated
arm whose material could be read either as a sorted cell population or as a
subject code; donors discordant in age; a differentiation stage declared on one
arm only; a control defined as free of dementia rather than free of the disease
under study. The rule is one-directional by design, and it is the same rule that
makes the weight below an upper bound. It does not mean every ambiguity was
resolved against the data: two contested comparisons were **not** counted,
because their labels do decide — a knockout genotype present identically on both
arms isolates the treatment correctly in that genotype, and a study whose agent
differs from the one the group is named for is already counted once, in the
coherence verdict.

The weight of a defective comparison is its share of the sum of inverse
variances within the group — the same weight the random-effects model uses for
pooling — and is therefore an **upper bound**: it attributes to a defect the
entire weight of the study involved, even when only one of that study's
comparisons is affected. In one of the case studies below this pushes the figure
above 80%, because the single study that dominates the pooling is among those
carrying a defect; the number is reported as it comes out.

Where this count was not performed, the summary card says so. It does not
report a weight without the count that gives it meaning.

### Declared limitations of this measure

- the proportion of groups containing at least one defective comparison **grows
  with group size**, so it is not a quality measure that can be compared across
  groups of different size;
- the weight column is an **upper bound**, as described above;
- a defect does not imply a wrong result. The agonist/antagonist pair among the
  case studies above contains defective comparisons in both groups and still
  produces the strongest agreement in this work.

### Provenance

Run identifier: `` `r lb$run_metadata$run_id` ``
Generated: `` `r lb$run_metadata$timestamp` ``
Package version: `` `r lb$run_metadata$package_version` ``

The `r nrow(lb$selection_resolved)` groups shown in this document, with their
internal identifiers for cross-checking against the pooled results:

```{r selection-table}
# SOLO le colonne pubblicabili. La colonna `notes` del file di selezione
# contiene appunti di lavoro -- in italiano, col gergo interno e con i numeri
# di un run precedente scritti a mano -- e finiva stampata per intero in questa
# appendice. Resta nel CSV accanto al documento, che e' il posto giusto per
# chi verifica; qui no.
COLONNE_SELEZIONE <- c(
  cluster_id    = "Group identifier",
  label_paper   = "Contrast",
  method        = "Pooling method",
  k_effective   = "Pooled studies",
  n_sig_FDR05   = "Significant genes"
)
sel_vista <- lb$selection_resolved[, intersect(names(COLONNE_SELEZIONE),
                                               names(lb$selection_resolved)),
                                   drop = FALSE]
names(sel_vista) <- unname(COLONNE_SELEZIONE[names(sel_vista)])
DT::datatable(sel_vista, rownames = FALSE, options = list(pageLength = 25))
```

Configuration used to generate the figures:

```{r config-print}
yaml::as.yaml(lb$run_metadata$config) |> cat()
```
