---
title: "P4 β — ARCHS4 human coverage report"
date: "`r Sys.Date()`"
output:
  quarto::html_document:
    toc: true
    theme: cosmo
---

# Pipeline coverage P4 β

```{r setup, include=FALSE}
library(jsonlite)
library(dplyr)
library(ggplot2)
recs1 <- stream_in(file("analysis/p4-output/<stage1-run-id>/predictions.jsonl"), verbose = FALSE)
recs2 <- stream_in(file("analysis/p4-output/<stage2-run-id>/predictions.jsonl"), verbose = FALSE)
multi <- read.table("analysis/p4-output/p4-beta-etl-multiseries.tsv",
                    sep = "\t", header = TRUE)
skipped <- read.table("analysis/p4-output/p4-beta-etl-skipped.tsv",
                      sep = "\t", header = TRUE)
```

## 1. ETL coverage

- Total sample in ARCHS4 v2.5 H5: TODO_FROM_LOG
- Sample includibili (filtri organism + library + length): `r nrow(recs1) + nrow(skipped)`
- Sample skippati per reason:
```{r}
table(skipped$skip_reason)
```

## 2. Series-id-resolver coverage

- Sample multi-series totali: `r nrow(multi)`
- Branch distribution:
```{r}
table(multi$resolver_branch)
```
- Tiebreak (`tiebreak_both_srp`) sample count: `r sum(multi$resolver_branch == "tiebreak_both_srp")`
- Fallback (`fallback_no_srp`) sample count: `r sum(multi$resolver_branch == "fallback_no_srp")`

## 3. Stage1 classification coverage

```{r}
schema_valid_rate <- mean(!is.na(recs1$primary_role))
cat(sprintf("Schema valid: %.2f%%\n", 100 * schema_valid_rate))
table(recs1$primary_role)
```

## 4. Stage2 study-level coverage

- Studi unici classificati: `r length(unique(recs2$series_id_resolved))`
- Distribuzione `design_role`:
```{r}
table(recs2$design_role)
```

## 5. Comparability_anchor distribution (top-20)

```{r}
recs2 |>
  count(comparability_anchor, sort = TRUE) |>
  slice_head(n = 20) |>
  knitr::kable()
```

## 6. Comparison vs P4 α

α stage1 cs50: 130.784 sample → 6649 record stage2 cs50.
β stage1: `r nrow(recs1)` sample → ? record stage2.

(da popolare con dati reali post-run)
